Criticisms of the IHME COVID-19 model

### 2024-07-02 text/20/ihme-model.Rmd
Updated 21/ihme-model

---
title: "Criticisms of the IHME COVID-19 model"
author: "Steve Simon"
source: new
date: 2020-04-23
categories:
- Blog post
tags:
- Epidemiology
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE)
suppressMessages(suppressWarnings(library(tidyverse)))
```

@@

```{r g1, fig.width=4, fig.height=2}
data.frame(
  x <- seq(-2.8, -1.5, length=100),
  y <- dnorm(x)) %>%
  ggplot(aes(x, y)) +
    geom_line() + 
    theme_void() + 
    expand_limits(x=c(-3, 0.5)) +
    expand_limits(y=c(-0.02, pnorm(0))) +
    geom_text(x=-2.5, y=-0.02, label="Time") +
    geom_text(x=-3, y=0.07, angle=90, label="Deaths")
```

@@

```{r g2, fig.width=4, fig.height=2}
data.frame(
  x <- seq(-2.8, -0.25, length=100),
  y <- dnorm(x)) %>%
  ggplot(aes(x, y)) +
    geom_line() + 
    theme_void() + 
    expand_limits(x=c(-3, 0.5)) +
    expand_limits(y=c(-0.02, pnorm(0))) +
    geom_text(x=-2.5, y=-0.02, label="Time") +
    geom_text(x=-3, y=0.07, angle=90, label="Deaths")
```

@@

```{r g3, fig.width=4, fig.height=2}
data.frame(
  x <- seq(-2.8, 0.5, length=100),
  y <- dnorm(x)) %>%
  ggplot(aes(x, y)) +
    geom_line() + 
    theme_void() + 
    expand_limits(x=c(-3, 0.5)) +
    expand_limits(y=c(-0.02, pnorm(0))) +
    geom_text(x=-2.5, y=-0.02, label="Time") +
    geom_text(x=-3, y=0.07, angle=90, label="Deaths")
```

@@

Murray CJL. Forecasting the impact of the first wave of the COVID-19 pandemic on hospital demand and deaths for the USA and European Economic Area countries. Preprint, 2020-04-21. Available in [pdf format](http://www.healthdata.org/sites/default/files/files/Projects/COVID/RA_COVID-forecasting-USA-EEA_042120.pdf).

@@

Murray CJL. Forecasting COVID-19 impact on hospital bed-days, ICU-days, ventilator-days and deaths by US state in the next 4 months. Preprint, 2020-03-26. Available in [pdf format](https://www.medrxiv.org/content/10.1101/2020.03.27.20043752v1.full.pdf).

@@

Xu B, Gutierrez B, Mekaru S, Sewalk K, Goodwin L, Loskill A, Cohn EL, Hswen Y, Hill SC, Cobo MM, Zarebski AE, Li S, Wu C-H, Hulland E, Morgan JD, Wang L, O'Brien K, Scarpino SV, Brownstein JS, Pybus OG, Pigott DM, Kraemer MUG. Epidemiological data from the COVID-19 outbreak, real-time case information. Scientific Data, 2020-03-24. doi:10.1038/s41597-020-0448-0. Available in [html format](https://www.nature.com/articles/s41597-020-0448-0) or [pdf format](https://www.nature.com/articles/s41597-020-0448-0.pdf).

@@

IHME COVID-19 model FAQs. Available in [html format](http://www.healthdata.org/covid/faqs).

@@

Aizenman N. 5 Key Facts Not Explained In White House COVID-19 Projections. National Public Radio 2020-04-01. Available in [html format](https://www.npr.org/sections/health-shots/2020/04/01/824744490/5-key-facts-the-white-house-isnt-saying-about-their-covid-19-projections).

@@

Bui Q, Katz J, Parlapiano A, and Sanger-Katz, M. What 5 Coronavirus Models Say the Next Month Will Look Like. The New York Times, 2020-04-22. Available in [html format](https://www.nytimes.com/interactive/2020/04/22/upshot/coronavirus-models.html).

@@

Chow D. What we know about the coronavirus model the White House unveiled. NBC News, 2020-04-01. Available in [html format](https://www.nbcnews.com/science/science-news/what-we-know-about-coronavirus-model-white-house-unveiled-n1173601).

@@

Druke N, Silver N. How One Modeler Is Trying To Forecast The Toll Of COVID-19. FiveThirtyEight 2020-04-22. Available in [video format](https://fivethirtyeight.com/features/politics-podcast-how-one-modeler-is-trying-to-forecast-the-toll-of-covid-19/) and as a [podcast](https://podcasts.apple.com/us/podcast/model-talk-forecasting-the-toll-of-covid-19/id1077418457).

@@

Begley S. Influential Covid-19 model uses flawed methods and shouldn?t guide U.S. policies, critics say. Stat 2020-04-17. Available in [html format](https://www.statnews.com/2020/04/17/influential-covid-19-model-uses-flawed-methods-shouldnt-guide-policies-critics-say/).

@@

Etzioni R, COVID-19 Models: What Makes Them Tick? Timmerman Report, 2020-04-07. Available in [html format](https://timmermanreport.com/2020/04/covid-19-models-what-makes-them-tick/).

@@

Mortality Modeling FAQ. Available in [html format](https://covid-19.tacc.utexas.edu/mortality-projection-faq/).

@@

Social Distancing: Data Models for a Model Response to an Outbreak. Available in [html format](https://sph.umich.edu/news/2020posts/data-models-for-a-model-response-to-an-outbreak.html).

@@

Robinson D. David's COVID Numbers. Available in [html format](https://davidscovidnumbers.wordpress.com/).

@@

Jewell NP, Lewnard JA, Jewell BL. Caution Warranted: Using the Institute for Health Metrics and Evaluation Model for Predicting the Course of the COVID-19 Pandemic. Ann Intern Med. 2020. Available in [html format](https://doi.org/10.7326/M20-1565).

@@

Chowell G, Sattenspiel L, Bansal S, Viboud C. Mathematical models to characterize early epidemic growth: A review. Phys Life Rev. 2016;18:66?97. doi:10.1016/j.plrev.2016.07.005. Available in [html format](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5348083/) or [pdf format](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5348083/pdf/nihms-802148.pdf).

@@

Kissler SM, Tediganto C, Goldstein E, Grad YH, Lipsitch M. Projecting the transmission dynamics of SARS-CoV-2 through the postpandemic period. Science 2020-04-14. DOI: 10.1126/science.abb5793, Available in [html format](https://science.sciencemag.org/content/early/2020/04/24/science.abb5793).

@@

FRED Modeling System. Available in [html format](https://fred.publichealth.pitt.edu/).

@@

Ayoub HH, Chemaitelly H, Mumtaz GR, Seedat S, Awad SF, Makhoul M, Abu-Raddad LJ. Characterizing key attributes of the epidemiology of COVID-19 in China: Model-based estimations. Preprint published in Medrxiv, 2020-04-11. Available in [pdf format](https://www.medrxiv.org/content/10.1101/2020.04.08.20058214v1.full.pdf).

@@

[US Predictions](https://mrc-ide.github.io/covid19usa/#/)

@@

Wang L, Wang G, Gao L, Li X, Yu S, Kim M, Wang Y, Gu Z. Spatiotemporal Dynamics, Nowcasting and Forecasting of COVID-19 in the United States. Preprint published in Arxiv, 2020-04-30. Available in [pdf format](https://arxiv.org/pdf/2004.14103.pdf).

+ This article fits an SEIR model to individual counties in the United States with a large number of county specific covariates. The model produces a [nice dashboard](https://covid19.stat.iastate.edu/).

@@

>>>>

---
title: "Criticisms of the IHME COVID-19 model"
author: "Steve Simon"
source: new
date: 2020-04-23
categories:
- Blog post
tags:
- Epidemiology
- Rmarkdown graphs
output: html_document
page_update: complete
---

```{r ihme-model-01, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE)
suppressMessages(suppressWarnings(library(tidyverse)))
```

@@

```{r ihme-model-02, fig.width=4, fig.height=2}
data.frame(
  x <- seq(-2.8, -1.5, length=100),
  y <- dnorm(x)) %>%
  ggplot(aes(x, y)) +
    geom_line() + 
    theme_void() + 
    expand_limits(x=c(-3, 0.5)) +
    expand_limits(y=c(-0.02, pnorm(0))) +
    geom_text(x=-2.5, y=-0.02, label="Time") +
    geom_text(x=-3, y=0.07, angle=90, label="Deaths")
```

@@

```{r ihme-model-03, fig.width=4, fig.height=2}
data.frame(
  x <- seq(-2.8, -0.25, length=100),
  y <- dnorm(x)) %>%
  ggplot(aes(x, y)) +
    geom_line() + 
    theme_void() + 
    expand_limits(x=c(-3, 0.5)) +
    expand_limits(y=c(-0.02, pnorm(0))) +
    geom_text(x=-2.5, y=-0.02, label="Time") +
    geom_text(x=-3, y=0.07, angle=90, label="Deaths")
```

@@

```{r ihme-model-04, fig.width=4, fig.height=2}
data.frame(
  x <- seq(-2.8, 0.5, length=100),
  y <- dnorm(x)) %>%
  ggplot(aes(x, y)) +
    geom_line() + 
    theme_void() + 
    expand_limits(x=c(-3, 0.5)) +
    expand_limits(y=c(-0.02, pnorm(0))) +
    geom_text(x=-2.5, y=-0.02, label="Time") +
    geom_text(x=-3, y=0.07, angle=90, label="Deaths")
```

@@

Murray CJL. Forecasting the impact of the first wave of the COVID-19 pandemic on hospital demand and deaths for the USA and European Economic Area countries. Preprint, 2020-04-21. Available in [pdf format][mur1].

@@

Murray CJL. Forecasting COVID-19 impact on hospital bed-days, ICU-days, ventilator-days and deaths by US state in the next 4 months. Preprint, 2020-03-26. Available in [pdf format][mur2].

@@

Xu B, Gutierrez B, Mekaru S, Sewalk K, Goodwin L, Loskill A, Cohn EL, Hswen Y, Hill SC, Cobo MM, Zarebski AE, Li S, Wu C-H, Hulland E, Morgan JD, Wang L, O'Brien K, Scarpino SV, Brownstein JS, Pybus OG, Pigott DM, Kraemer MUG. Epidemiological data from the COVID-19 outbreak, real-time case information. Scientific Data, 2020-03-24. doi:10.1038/s41597-020-0448-0. Available in [html format][xub1] or [pdf format][xub2].

@@

IHME COVID-19 model FAQs. Available in [html format][ihm1].

@@

Aizenman N. 5 Key Facts Not Explained In White House COVID-19 Projections. National Public Radio 2020-04-01. Available in [html format][aiz1].

@@

Bui Q, Katz J, Parlapiano A, and Sanger-Katz, M. What 5 Coronavirus Models Say the Next Month Will Look Like. The New York Times, 2020-04-22. Available in [html format][bui1].

@@

Chow D. What we know about the coronavirus model the White House unveiled. NBC News, 2020-04-01. Available in [html format][cho1].

@@

Druke N, Silver N. How One Modeler Is Trying To Forecast The Toll Of COVID-19. FiveThirtyEight 2020-04-22. Available in [video format][dru1] and as a [podcast][dru2].

@@

Begley S. Influential Covid-19 model uses flawed methods and shouldn?t guide U.S. policies, critics say. Stat 2020-04-17. Available in [html format][beg1].

@@

Etzioni R, COVID-19 Models: What Makes Them Tick? Timmerman Report, 2020-04-07. Available in [html format][etz1].

@@

Mortality Modeling FAQ. Available in [html format][mor1].

@@

Social Distancing: Data Models for a Model Response to an Outbreak. Available in [html format][soc1].

@@

Robinson D. David's COVID Numbers. Available in [html format][rob1].

@@

Jewell NP, Lewnard JA, Jewell BL. Caution Warranted: Using the Institute for Health Metrics and Evaluation Model for Predicting the Course of the COVID-19 Pandemic. Ann Intern Med. 2020. Available in [html format][jew1].

@@

Chowell G, Sattenspiel L, Bansal S, Viboud C. Mathematical models to characterize early epidemic growth: A review. Phys Life Rev. 2016;18:66?97. doi:10.1016/j.plrev.2016.07.005. Available in [html format][cho2] or [pdf format][cho3].

@@

Kissler SM, Tediganto C, Goldstein E, Grad YH, Lipsitch M. Projecting the transmission dynamics of SARS-CoV-2 through the postpandemic period. Science 2020-04-14. DOI: 10.1126/science.abb5793, Available in [html format][kis1].

@@

FRED Modeling System. Available in [html format][fre1].

@@

Ayoub HH, Chemaitelly H, Mumtaz GR, Seedat S, Awad SF, Makhoul M, Abu-Raddad LJ. Characterizing key attributes of the epidemiology of COVID-19 in China: Model-based estimations. Preprint published in Medrxiv, 2020-04-11. Available in [pdf format][aya1].

@@

[US Predictions][usp1]

@@

Wang L, Wang G, Gao L, Li X, Yu S, Kim M, Wang Y, Gu Z. Spatiotemporal Dynamics, Nowcasting and Forecasting of COVID-19 in the United States. Preprint published in Arxiv, 2020-04-30. Available in [pdf format][wan1].

+ This article fits an SEIR model to individual counties in the United States with a large number of county specific covariates. The model produces a [nice dashboard][ias1].

@@

[aiz1]: https://www.npr.org/sections/health-shots/2020/04/01/824744490/5-key-facts-the-white-house-isnt-saying-about-their-covid-19-projections
[aya1]: https://www.medrxiv.org/content/10.1101/2020.04.08.20058214v1.full.pdf
[beg1]: https://www.statnews.com/2020/04/17/influential-covid-19-model-uses-flawed-methods-shouldnt-guide-policies-critics-say/
[bui1]: https://www.nytimes.com/interactive/2020/04/22/upshot/coronavirus-models.html
[cho1]: https://www.nbcnews.com/science/science-news/what-we-know-about-coronavirus-model-white-house-unveiled-n1173601
[cho2]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5348083/
[cho3]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5348083/pdf/nihms-802148.pdf
[dru1]: https://fivethirtyeight.com/features/politics-podcast-how-one-modeler-is-trying-to-forecast-the-toll-of-covid-19/
[dru2]: https://podcasts.apple.com/us/podcast/model-talk-forecasting-the-toll-of-covid-19/id1077418457
[etz1]: https://timmermanreport.com/2020/04/covid-19-models-what-makes-them-tick/
[fre1]: https://fred.publichealth.pitt.edu/
[ias1]: https://covid19.stat.iastate.edu/
[ihm1]: http://www.healthdata.org/covid/faqs
[jew1]: https://doi.org/10.7326/M20-1565
[kis1]: https://science.sciencemag.org/content/early/2020/04/24/science.abb5793
[mor1]: https://covid-19.tacc.utexas.edu/mortality-projection-faq/
[mur1]: http://www.healthdata.org/sites/default/files/files/Projects/COVID/RA_COVID-forecasting-USA-EEA_042120.pdf
[mur2]: https://www.medrxiv.org/content/10.1101/2020.03.27.20043752v1.full.pdf
[rob1]: https://davidscovidnumbers.wordpress.com/
[soc1]: https://sph.umich.edu/news/2020posts/data-models-for-a-model-response-to-an-outbreak.html
[usp1]: https://mrc-ide.github.io/covid19usa/#/
[wan1]: https://arxiv.org/pdf/2004.14103.pdf
[xub1]: https://www.nature.com/articles/s41597-020-0448-0
[xub2]: https://www.nature.com/articles/s41597-020-0448-0.pdf
